Handling String-Number Conversion in String Constraint Solving

### 2019-11-15 draft.tex
address the comments from Lukas

We pick a PFA for each variable in $\vars$ at the beginning of the decision procedure and then construct the set of linear integer constraints in a stepwise manner. In each step, we take a string constraint from $\Phi_I$ and depending on the type of the string constraint, we execute the corresponding procedure to produce the linear integer constraints and add them to the output linear formulas $\Phi_O$.

@@

Assume that the constraint we take from $\Phi_I$ is a string-number conversion constraint  $\phi_s \defeq n=\sti{x}$. We will discuss how to make the corresponding linear formulas in this section.

Let us begin the discussion with a simple example, assume that we use the PFA in Figure~\ref{fig:sfa} (a) to restrict the domain of $x$. Then we know that when $0\leq v_0,v_1 \leq 9$, then $n$ is a positive integer value, and otherwise $n =-1$. So we should first add the constraint $ ((0\leq v_0\leq 9) \wedge (0\leq v_1\leq 9)) \vee n=-1 $.

@@

\paragraph{PFA that allows efficient string-number conversion: } In the previous discussion, we know that it is unlikely to use arbitrary PFA and gain efficient solving procedure, due to the requirement to solve exponential integer constraints. The problem is that each loop in the PFA creates an exponential component. An immediate idea would be that maybe we can restrict the variable to some specific PFA that allow more efficient constraint solving. At the same time, we want to keep some good property, i.e., the search space covers all integers with at most $m$-digits.

Our first attempt is a PFA with no loop, i.e., a straight line structure. The corresponding integer constraint does not have exponential components. However, it does not ensure the property that all numbers with at most $m$-digits are covered. Consider the example $\sti{x}=10 \wedge |x|=5$. The number $10$ has only $2$-digits, at the first glance, a straight-line PFA with two transitions, i.e., the PFA \scalebox{0.5}{\tikzset{state/.style={circle,draw=blue!50,fill=blue!20,
			thick,inner sep=0pt,minimum size=6mm}, initial text=$ $}
	\begin{tikzpicture} 
	\node[state,initial] (q0)  {$q_0$};
	\node[state] (q1) [right = 1cm of q0] {$q_1$};
	\node[state,accepting] (q2) [right = 1cm of q1] {$q_2$};

	\draw[->] (q0) edge node [below]{$v_1$} (q1) ;
	\draw[->] (q1) edge node [below]{$v_2$} (q2) ;
	\end{tikzpicture}} should be sufficient for the domain restriction of $x$. If we do so, we will conclude that the formula is unsatisfiable, because the length of $x$ cannot be $5$  under this domain restriction.

However, the formula is satisfiable when $x=``00010"$. Observe that $\sti{``00010"}=10$. The key is that even for a bounded integer, the corresponding string can be of unbounded length with arbitrarily many `$0$' characters at the front. In order to handle all number with at most $m$ digits, we propose to restrict the domain of $x$ to a PFA in Figure~\ref{fig:sfa_its}.  The PFA has a self-loop transition labeled $v_0$ at $q_0$ and we add a constraint $v_0=0$ to $\Phi_O$ for such PFA. With this transition, we can ensure that the under-approximation handles all numbers with at most $m$ digits. Although this PFA still has a loop, but it will not create any exponential component because the loop only represents a sequence of `0' that occurs in the prefix of $x$. It will not affect the integer value of $n=\sti{x}$.

@@

	\caption{An example of a symbolic flat automaton}
	\label{fig:sfa_its}
\end{figure}

Such PFA handles bounded integers (e.g., 10-digits or 64-bits) precisely and under-approximate the possible values of unbounded integers. It works very efficiently because we just need to use linear constraints to characterize the relation between $x$ and $n$. 
Remember that the character variables can be assigned $\epsilon$. In order to simplify the construction, we an encoding to shift all the $\epsilon$-transitions to the least significant digits, i.e., if $v_j = \epsilon = -1$ then $v_k = -1$ for all $k\in [j+1,m]$.  To explain this encoding, we first define two macro function:

@@

The correctness of this procedure is expressed in Lemma~\ref{lemma:sticorrect_pos} and Lemma~\ref{lemma:sticorrect_neg} below. We separate the lemmas to cases where $n\neq -1$ and $n=-1$. 
\yfc{The correctness lemmas are still buggy}
%
%\newcommand\restrict[1]{\raisebox{-.5ex}{$\big|$}_{#1}}
\begin{lemma}\label{lemma:sticorrect_pos}
	$\semof {\Psi_n \wedge \parikhfof {A_x} \wedge (v_0 = 0)\wedge (n\neq -1) }_{V_x \cup \#V_x \cup \{n\}} = \{(I\cup\pim \cup I_n) \mid  I_n(n)= \sti{\decode {A_x}(I\cup\pim)}\} $. 
\end{lemma}
In this case, $x$ encodes a number. The lemma says when the models we add to $\Phi_O$ represents a word in $\semof{A_x}$ whose integer value is exactly the model for $n$.

\begin{lemma}\label{lemma:sticorrect_neg}
	$\semof {\Psi_n \wedge \parikhfof {A_x} \wedge (v_0 = 0)\wedge (n = -1) }_{V_x \cup \#V_x \cup \{n\}} = 
	\{(I\cup\pim \cup I_n) \mid  I_n(n)= -1 \wedge \decode {A_x}(I\cup\pim) \in [0,9]^+  \}$
\end{lemma}
In this case, $x$ does not encode a number. The lemma says  the models we add to $\Phi_O$ represents words in both $\semof{A_x}$ and $[0,9]^+$, which means not a number, and $n$ is $-1$.

>>>>

We pick a PFA for each variable in $\vars$ at the beginning of the decision procedure and then construct the set of linear integer constraints in a stepwise manner. In each step, we take a string constraint from $\Phi_I$ and depending on the type of the string constraint, we execute the corresponding procedure to produce the linear integer constraints and add them to the output linear formulas $\Phi_O$. We also add constraints to $\Phi_O$ to ensure all character variables of the PFAs only pick values from $\Sigma_\epsilon$.

@@

Assume that the constraint we take from $\Phi_I$ is a string-number conversion constraint  $\phi_s \defeq n=\sti{x}$. In this section, we how to make the corresponding linear formulas for $\Phi_O$.

We begin the discussion with a simple example, assume that we use the PFA in Figure~\ref{fig:sfa} (a) to restrict the domain of $x$. Then we know that when $0\leq v_0,v_1 \leq 9$, then $n$ is a positive integer value, and otherwise $n =-1$. So we should first add the constraint $ ((0\leq v_0\leq 9) \wedge (0\leq v_1\leq 9)) \vee (n=-1 \wedge (v_0 >9 \vee v_1 >9))$.

@@

\paragraph{PFA that allows efficient string-number conversion: } Now we know that it is unlikely to use arbitrary PFA and gain efficient solving procedure, due to the requirement to solve exponential integer constraints. The cause of the problem is that each loop in the PFA creates an exponential component in the constraint. An immediate idea would be that maybe we can pick some specific PFA that allow more efficient constraint solving. At the same time, we want to keep a good property, that is, the search space covers all integers up to $m$-digits.

Our first attempt is a PFA without loop, i.e., a straight line structure 
\scalebox{0.5}{\tikzset{state/.style={circle,draw=blue!50,fill=blue!20,
			thick,inner sep=0pt,minimum size=6mm}, initial text=$ $}
	\begin{tikzpicture} 
		\node[state,initial] (q0)  {$q_0$};
		\node[state] (q1) [right = 1cm of q0] {$q_1$};
		\node (q2) [right = 0.8cm of q1]{};
		\node (q3) [right = 1cm of q2]{};
		\node[state,accepting] (qm) [right = 1cm of q3] {$q_m$};

		\draw[->] (q0) edge node [below]{$v_1$} (q1) ;
		\draw[->] (q1) edge node [below]{$v_2$} (q2) ;
		\draw[dotted] (q2) edge (q3) ;
		\draw[->] (q3) edge node [below]{$v_m$} (qm) ;
	\end{tikzpicture}}
. The corresponding integer constraint does not have exponential components. However, it does cover all numbers with at most $m$-digits. Consider the example $\sti{x}=10 \wedge |x|=5$. The number $10$ has only $2$-digits, at the first glance, a straight-line PFA with two transitions, i.e., the PFA
    \scalebox{0.5}{\tikzset{state/.style={circle,draw=blue!50,fill=blue!20,
			thick,inner sep=0pt,minimum size=6mm}, initial text=$ $}
	\begin{tikzpicture} 
	\node[state,initial] (q0)  {$q_0$};
	\node[state] (q1) [right = 1cm of q0] {$q_1$};
	\node[state,accepting] (q2) [right = 1cm of q1] {$q_2$};

	\draw[->] (q0) edge node [below]{$v_1$} (q1) ;
	\draw[->] (q1) edge node [below]{$v_2$} (q2) ;
	\end{tikzpicture}
	} should be sufficient for the domain restriction of $x$. If we do so, we will conclude that the formula is unsatisfiable, because the length of $x$ cannot be $5$  under this domain restriction.

However, the formula is satisfiable when $x=``00010"$. Observe that $\sti{``00010"}=10$. The key is that even for a bounded integer, the corresponding string can be of unbounded length with arbitrarily many `$0$' characters at the front. In order to handle all numbers with up to $m$-digits, we propose to restrict the domain of $x$ to a PFA in Figure~\ref{fig:sfa_its}.  The PFA has a self-loop transition labeled $v_0$ at $q_0$ and we add a constraint $\Psi_{v_0}\defeq v_0=0$ to $\Phi_O$ for such PFA. With this transition, we can ensure that the under-approximation handles all numbers with at most $m$ digits. Although this PFA still has a loop, it will not create any exponential component because the loop only represents a sequence of ``0'' at the front of $x$. It will not affect the integer value of $n=\sti{x}$.

@@

	\caption{PFA for string-number conversion constraints}
	\label{fig:sfa_its}
\end{figure}

Such PFA handles bounded integers (e.g., 10-digits or 64-bits) precisely and under-approximate the possible values of unbounded integers. It works very efficiently because we just need to use linear constraints to characterize the relation between $x$ and $n$. 
Remember that the character variables can be assigned $\epsilon$. In order to simplify the construction, we use an encoding to shift all the $\epsilon$-transitions to the least significant digits, i.e., if $v_j = \epsilon = -1$ then $v_k = -1$ for all $k\in [j+1,m]$.  To explain this encoding, we first define two macro function:

@@

The correctness of this procedure is expressed in Lemma~\ref{lemma:sticorrect} below. We separate the lemmas to cases where $n\neq -1$ and $n=-1$.

%
%\newcommand\restrict[1]{\raisebox{-.5ex}{$\big|$}_{#1}}

\begin{lemma}\label{lemma:sticorrect}

	The interpretation $(I\cup\pim \cup I_n)  \in \semof {\Psi_n \wedge \Psi_{v_0}\wedge \parikhfof {A_x} }_{V_x \cup \#V_x \cup \{n\}} $ if and only if 
	\begin{center}
	 $\left(\begin{array}{c} 
	 I_n(n)= \sti{\decode {A_x}(I\cup\pim)} \ \wedge \\
	 I(v_i)=-1 \wedge i<j \Longrightarrow  I(v_j)=-1
	 \end{array}\right)$\\ or\\
	 $\left(\begin{array}{c} 
	 I_n(n)= -1\ \wedge \decode {A_x}(I\cup\pim) \not\in [0,9]^+
	 \end{array}\right)$
	\end{center}

\end{lemma}
\yfc{sorry still ugly, the shifting of epsilon makes it a mess here}

\hide{
\begin{lemma}\label{lemma:sticorrect_pos}
	$\semof {\Psi_n \wedge \Psi_{v_0}\wedge \parikhfof {A_x} }_{V_x \cup \#V_x \cup \{n\}} = \{(I\cup\pim \cup I_n) \mid  I_n(n)= \sti{\decode {A_x}(I\cup\pim)}\} $. 
\end{lemma}
In this case, $x$ encodes a number. The lemma says when the models we add to $\Phi_O$ represents a word in $\semof{A_x}$ whose integer value is exactly the model for $n$.

\begin{lemma}\label{lemma:sticorrect_neg}
	$\semof {\Psi_n \wedge \parikhfof {A_x} \wedge (v_0 = 0)\wedge (n = -1) }_{V_x \cup \#V_x \cup \{n\}} = 
	\{(I\cup\pim \cup I_n) \mid  I_n(n)= -1 \wedge \decode {A_x}(I\cup\pim) \not\in [0,9]^+  \}$
\end{lemma}
In this case, $x$ does not encode a number. The lemma says  the models we add to $\Phi_O$ represents words in both $\semof{A_x}$ and $[0,9]^+$, which means not a number, and $n$ is $-1$.}
